Minimax-optimal Laplacian Eigenmaps regression over Sobolev Spaces with Neighborhood Graphs

### 2021-04-28 manuscripts/laplacian_eigenmaps/laplacian_eigenmaps_regression.tex
Completed the proof of Proposition 2 in Laplacian Eigenmaps paper.
Read through Laplacian Eigenmaps paper for errors. Ready to be merged.

%%% Vectors
\newcommand{\thetast}{\theta^{\star}}
\newcommand{\betap}{\beta^{(p)}}
\newcommand{\betaq}{\beta^{(q)}}
\newcommand{\vardeltapq}{\varDelta^{(p,q)}}
\newcommand{\lambdavec}{\boldsymbol{\lambda}}

@@

Suppose we observe data $X_1,\ldots,X_n \in \Reals^d$, sampled independently from an unknown distribution $P$. As a replacement for $P$, we form a geometric graph $G$ over the observed data, with vertices at $X_1,\ldots,X_n$ and weighted edges $W_{ij}$ corresponding to proximity between samples $X_i$ and $X_j$. Geometric graphs encode both local information and global information about $P$ in an extremely general manner. For this reason they have been leveraged to conduct many different fundamental statistical tasks, such clustering, manifold learning, semi-supervised learning, classification, and regression \textcolor{red}{(TODO): references}. Substantial theoretical progress has been made regarding the consistency of \emph{graph-based learning}, that is, learning algorithms defined with respect to geometric graphs \textcolor{red}{(TODO): references}. This work sheds light on why such procedures work, by showing that they converge to interesting \emph{continuum} limits as $n \to \infty$. However, thus far little has been said regarding the optimality of graph-based learning methods, even for classic statistical tasks.

In this paper we focus on the theoretical statistical properties of a particularly popular graph-based learning method for regression. We assume that in addition to the design points $X_1,\ldots,X_n$ one observes real-valued responses $Y_1,\ldots,Y_n$, and seeks to learn the unknown regression function $f_0(x) := E[Y|X = x]$. The specific graph-based method we study is \emph{Laplacian eigenmaps}, first introduced by \cite{belkin03a}, which projects the response vector ${\bf Y} = (Y_1,\ldots,Y_n)$ onto the span of eigenvectors of a graph Laplacian. We focus on the the unnormalized graph Laplacian $L$, which is a difference operator acting on functions $f: \{X_1,\ldots,X_n\} \to \Reals$ as follows,
\begin{equation}
\label{eqn:graph_laplacian}
Lf(X_i) = \sum_{j = 1}^{n} \bigl(f(X_i) - f(X_j)\bigr)W_{ij}. 
\end{equation}
The Laplacian $L$ can be seen as a discretization of a weighted Laplace-Beltrami operator $\Delta_P$, defined when $P$ admits a density $p$ as
\begin{equation}
\label{eqn:fokker_planck_1}
\Delta_Pf= -\frac{1}{p} \mathrm{div}(p^2 \nabla f).
\end{equation}
The eigenvectors of $L$ form an orthonormal basis of $L^2(P_n)$, and are approximations to the eigenfunctions of $\Delta_P$. Their corresponding eigenvalues are likewise estimates of eigenvalues of $\Delta_P$, and thus give a notion of smoothness to each eigenvector---roughly speaking, the smaller the eigenvalue, the smoother the corresponding eigenvector.

Laplacian eigenmaps is thus a type \emph{spectral series} method. Spectral series methods, and more general orthogonal series methods, are some of most classical methods for non-parametric regression. Traditionally, orthogonal series methods are defined with respect to a fixed set of basis functions, orthogonalized in $L^2(Q)$ with respect to some pre-determined reference measure $Q$. Spectral series estimators take a particular orthogonal basis, the eigenfunctions of $\Delta_Q$. In contrast, eigenvectors of the graph Laplacian are data-dependent objects, and adapt to the geometry of the unknown design distribution $P$ in a rich manner.

The theoretical properties of (classical) spectral series methods are various, and are in general extremely well-understood \textcolor{red}{(TODO): references}. In particular, they are minimax rate optimal for nonparametric regression over H\"{o}lder and Sobolev spaces. However, these methods suffer from some serious drawbacks. As a practical matter, even computing the eigenfunctions of $\Delta_Q$ is in general non-trivial. Moreover, the statistical optimality of such estimators also requires that the reference measure $Q$ be equal to $P$, whereas in this work we shall always assume $P$ is unknown. Of course there exist workarounds for these issues, for instance non-parametric least squares or \textcolor{red}{(...)}. We will comment more on these alternatives later; for now we merely emphasize that they constitute a fundamental change to the method under consideration. \textcolor{red}{Last sentence is weak, need to clarify, can maybe discuss with Ryan + Siva.}

In contrast, Laplacian eigenmaps seeks to directly approximate a orthogonal series method that uses the eigenfunctions of $\Delta_P$, as we elaborate on momentarily. Importantly, it avoids the (arguably unrealistic) assumption that $P$ and the eigenfunctions of $\Delta_P$ be known, by using the graph Laplacian eigenvectors as a stand-in. On the other hand its theoretical statistical properties, and in particular its optimality as a method for regression, remain poorly understood.

@@

Broadly seeking, our answer is yes. We consider two different data models, one of which assumes a full-dimensional, flat Euclidean domain, and the other of which assumes that the distribution $P$ is supported on a manifold of small intrinsic dimension $m$. We show that when the regression function $f_0$ is smooth, in the Sobolev sense of having weak derivatives bounded in $\Leb^2$ norm, Laplacian eigenmaps methods are statistically minimax optimal, for both estimation and goodness-of-fit testing. Our statements holds for different relations between the dimension $d$ (or $m$) and number of derivatives $s$, depending on the problem (estimation or testing), and are summarized in Tables~\ref{tbl:estimation_rates} and~\ref{tbl:testing_rates}.
\begin{table}
	\begin{center}
		\begin{tabular}{p{.2\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Smoothness order & Flat Euclidean (Model~\ref{def:model_flat_euclidean}) & Manifold (Model~\ref{def:model_manifold}) \\
			\hline
			$s \leq 3$ & $\bm{n^{-2s/(2s + d)}}$ & $\bm{n^{-2s/(2s + m)}}$ \\
			$s > 3$  & $\bm{n^{-2s/(2s + d)}}$ & $n^{-6/(6 + m)}$
		\end{tabular}
	\end{center}
	\caption{Summary of estimation rates over Sobolev balls. Bold font marks minimax optimal rates. In each case, rates hold for all $d \in \mathbb{N}$ (under Model~\ref{def:model_flat_euclidean}), and for all $m \in \mathbb{N}, 1 < m < d$ (under Model~\ref{def:model_manifold}). Although we suppress it for simplicity, in all cases when the Laplacian eigenmaps estimator is optimal, the dependence of the error rate on the radius $M$ of the Sobolev ball is also optimal, as long as $M \gtrsim n^{-1/2}$.}
	\label{tbl:estimation_rates}
\end{table}

\begin{table}
	\begin{center}
		\begin{tabular}{p{.175\textwidth} p{.175\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Smoothness order & Dimension & Flat Euclidean (Model~\ref{def:model_flat_euclidean}) & Manifold (Model~\ref{def:model_manifold}) \\
			\hline
			\multirow{2}{*}{$s = 1$} & $\dim(\mc{X}) < 4$ & $\bm{n^{-4s/(4s + d)}}$ & $\bm{n^{-4s/(4s + m)}}$ \\
			& $\dim(\mc{X}) \geq 4$ & $\bm{n^{-1/2}}$ & $\bm{n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s = 2$ or $3$} & $\dim(\mc{X}) \leq 4$  & $\bm{n^{-4s/(4s + d)}}$ & $\bm{n^{-4s/(4s + m)}}$ \\
			& $4 <\dim(\mc{X}) < 4s$  & $n^{-2s/(2(s - 1) + d)}$ & $n^{-2s/(2(s - 1) + m)}$\\
			& $\dim(\mc{X}) \geq 4s$ & $\bm{n^{-1/2}}$ & $\bm{n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s > 3$} & $\dim(\mc{X}) \leq 4$ & $\bm{n^{-4s/(4s + d)}}$ & $n^{-12/(12 + d)}$ \\
			& $4 < \dim(\mc{X}) < 4s$ & $n^{-2s/(2(s - 1) + d)}$ & $n^{-6/(4 + m)}$ \\
			& $\dim(\mc{X}) \geq 4s$ & $\bm{n^{-1/2}}$ & $\bm{n^{-1/2}}$ \\
		\end{tabular}
	\end{center}
	\caption{Summary of Laplacian eigenmaps testing rates over Sobolev balls. Bold font marks minimax optimal rates. Rates when $d > 4s$ assume that $f_0 \in L^4(P,M)$. Although we suppress it for simplicity, in all cases when the Laplacian eigenmaps test is optimal, the dependence of the error rate on the radius $M$ of the Sobolev ball is also optimal, as long as $M \gtrsim n^{-1/2}$.}
	\label{tbl:testing_rates}
\end{table}

\paragraph{Related Work.}
\textcolor{red}{(TODO)}: Flesh this out.
\begin{itemize}
	\item Summarize what is known on fixed graphs, drawing from our paper on Laplacian smoothing.
	\item \textcolor{red}{von Luxburg, Belkin, Slepcev} show that for a fixed index $k$, the eigenvector $v_k$ converges to a eigenfunction $\psi_k$ of the Laplace-Beltrami operator $\Delta_P$. \textcolor{red}{Shi, Burago, Garcia Trillos, Calder, Cheng} build on this, giving finite sample bounds, rates of convergence, and making statements uniform over $1 \leq k \leq k_{\max}$. This makes it clear that Laplacian eigenmaps is approximating a classical spectral series method, and may be a good alternative when the latter cannot be implemented (e.g. because $P$ or $\mc{X}$ are unknown.) It is tempting to try to directly use these results along with the triangle inequality to deduce that if (i)  spectral series methods are optimal, and (ii) Laplacian eigenmaps is converging to a spectral series method, then (iii) Laplacian eigenmaps must be also optimal. Unfortunately, this won't work, both because we will need to take an increasing number of eigenvectors (increasing, that is, with the sample size $n$) to achieve an optimal estimator and test, and because the best known rates of convergence of $v_k \to \psi_k$ are slower than the minimax rates.
	\item There has been limited analysis of the estimator $\wh{f}$. \cite{zhou2011} consider $\wh{f}$, but in the semi-supervised setting where the number of unlabeled points grows to infinity. In this case $\wh{f}$ reduces to a classical spectral series estimator, and there is no error incurred by approximating $\Delta_P$. \cite{lee2016} consider the \emph{diffusion maps} estimator---which uses the eigenvectors of a different normalization of the Laplacian $L$---in both the supervised and semi-supervised setups, but they derive suboptimal rates for the supervised problem. As far as we know, there has been no analysis of the test statistic $\wh{T}$ in the random design framework which we study.
	\item Alternatively, \cite{trillos2020} \textcolor{blue}{(Green 2021)} use the graph Laplacian operator to induce a penalty over functions $f: \{\bf X\} \to \Reals$. The \emph{Laplacian smoothing} estimator $\wh{f}_{\mathrm{LS}}$ is obtained by minimizing the sum of this penalty along with a data-fidelity term,
	\begin{equation*}
	\wh{f}_{\mathrm{LS}} := \|Y - f\|_n^2 + \lambda \dotp{L_{n,\varepsilon}f}{f}_n.
	\end{equation*}
	\textcolor{blue}{(Green 2021)} show that the resulting estimator is minimax optimal, but only for $s = 1$ and $d \leq 4$. In contrast, the Laplacian eigenmaps estimator is optimal for all $s$ and $d$. 
\end{itemize}

@@

We now outline the structure of the rest of this paper. In Section~\ref{sec:setup_main_results}, we formally define the regression problem and estimator we consider, and give some background about minimax regression over Sobolev spaces. In Sections~\ref{sec:minimax_optimal_laplacian_eigenmaps} and~\ref{sec:manifold_adaptivity}, we establish our main results regarding the optimality of Laplacian eigenmaps. The Laplacian eigenmaps estimator $\wh{f}$ is defined only at the design points $X_1,\ldots,X_n$. In Section~\ref{sec:out_of_sample}, we propose a method for out-of-sample extension of $\wh{f}$, and show that it has optimal out-of-sample error. In Section~\ref{sec:experiments} we examine the empirical behavior of Laplacian eigenmaps, and compare it to some natural competitors for non-parametric regression. We conclude with some discussion in Section~\ref{sec:discussion}.

\paragraph{Notation.}

\textcolor{red}{TODO}:

\begin{itemize}
	\item Define Lebesgue measure $\nu$. 
	\item Define Euclidean norm $\|\cdot\|$.
	\item Define $\Leb^q(\mc{X}), \Leb^2(P)$ and $\Leb^2(P_n)$. Define $C^k(\mc{X})$, and its norm. Define $C_0^k(\mc{X})$.
	\item Define asymptotic notation: $\ll$, $\lesssim$ and $\asymp$.
	\item Define min and max notation: $\vee$ and $\wedge$.
\end{itemize}

@@

In this section, we begin by giving a precise definition of our framework, and the Laplacian eigenmaps methods we study. We then review minimax rates for non-parametric regression over Sobolev spaces.

\subsection{Non-parametric regression with random design}
\label{sec:regression_laplacian_eigenmaps}

We will operate in the usual setting of non-parametric regression with random design, in which we observe independent random samples $(X_1,Y_1),\ldots,(X_n,Y_n)$. The design points $X_1,\ldots,X_n$ are sampled from a distribution $P$ with support $\mc{X} \subseteq \Rd$, and the responses follow
\begin{equation}
\label{eqn:model}
Y_i = f_0(X_i) + w_i,
\end{equation}
with regression function $f_0: \mc{X} \to \Reals$, and $w_i \sim N(0,1)$ independent Gaussian noise. For simplicity we will assume throughout that the noise has unit-variance, but all of our results extend in a straightforward manner to the case where the variance is equal to a known positive value.

@@

\begin{definition}[Flat Euclidean model]
	\label{def:model_flat_euclidean}
	The data $(X_1,Y_1),\ldots,(X_n,Y_n)$ are sampled according to~\eqref{eqn:model}. The support $\mc{X}$ of the design distribution $P$ is an open, connected, and bounded subset of $\Rd$, with Lipschitz boundary. The distribution $P$ admits a density $p$ with respect to the $d$-dimensional Lebesgue measure $\nu$, which is bounded away from $0$ and $\infty$,
	\begin{equation*}
	0 < p_{\min} \leq p(x) \leq p_{\max} < \infty, \quad \textrm{for all $x \in \mc{X}$.}
	\end{equation*}
\end{definition}

>>>>

%%% Vectors
\newcommand{\thetast}{\theta^{\star}}
\newcommand{\betap}{\beta^{(p)}}
\newcommand{\betaq}{\beta^{(q)}}
\newcommand{\vardeltapq}{\varDelta^{(p,q)}}
\newcommand{\lambdavec}{\boldsymbol{\lambda}}
\newcommand{\bj}{{\bf j}}

@@

Suppose we observe data $X_1,\ldots,X_n \in \Reals^d$, sampled independently from an unknown distribution $P$. As a replacement for $P$, we form a geometric graph $G$ over the observed data, with vertices at $X_1,\ldots,X_n$ and weighted edges $W_{ij}$ corresponding to proximity between samples $X_i$ and $X_j$. Geometric graphs encode both local information and global information about $P$ in an extremely general manner. For this reason they have been leveraged to conduct many different fundamental statistical tasks, such as clustering, manifold learning, semi-supervised learning, classification, and regression \textcolor{red}{(TODO): references}. Substantial theoretical progress has been made regarding the consistency of \emph{graph-based learning}, that is, learning algorithms defined with respect to geometric graphs \textcolor{red}{(TODO): references}. This work sheds light on why such procedures work, by showing that they converge to interesting \emph{continuum} limits as $n \to \infty$. However, thus far little has been said regarding the optimality of graph-based learning methods, even for classic statistical tasks.

In this paper we focus on the theoretical statistical properties of a particularly popular graph-based learning method for regression. We assume that in addition to the design points $X_1,\ldots,X_n$ one observes real-valued responses $Y_1,\ldots,Y_n$, and seeks to learn the unknown regression function $f_0(x) := E[Y|X = x]$. The specific graph-based method we study is \emph{Laplacian eigenmaps}, first introduced by \cite{belkin03a}, which projects the response vector ${\bf Y} = (Y_1,\ldots,Y_n)$ onto the span of eigenvectors of a graph Laplacian. We focus on the the unnormalized graph Laplacian $L$, which is a difference operator acting on vectors $u \in \Reals^n$ as follows,
\begin{equation}
\label{eqn:graph_laplacian}
(Lf)_i = \sum_{j = 1}^{n} (u_i - u_j)W_{ij}. 
\end{equation}
The graph Laplacian $L$ is a discrete approximation to a weighted continuum Laplacian operator $\Delta_P$, defined when $P$ admits a differentiable density $p$ as
\begin{equation}
\label{eqn:fokker_planck_1}
\Delta_Pf= -\frac{1}{p} \mathrm{div}(p^2 \nabla f).
\end{equation}
The eigenvectors of $L$ form an orthonormal basis of $\Reals^n$, and serve as estimates of eigenfunctions of $\Delta_P$. Their corresponding eigenvalues are likewise estimates of eigenvalues of $\Delta_P$, and give a notion of smoothness to each eigenvector---roughly speaking, the smaller the eigenvalue, the smoother the corresponding eigenvector.

We can therefore view Laplacian eigenmaps as a twist on a very classical approach to nonparametric regression: \emph{spectral projection}, or more generally \emph{orthogonal series}, regression. Classically, in orthogonal series regression one fixes a reference measure $Q$, takes an ordered orthonormal basis $\psi_1,\psi_2,\ldots$ of $L^2(Q)$, computes empirical Fourier coefficients $\dotp{{\bf Y}}{\psi_k}_n$, and uses the first few terms in the resulting Fourier series to construct an estimator. Regression by spectral projection is a special case of this general setup, where one takes $\psi_k$ to be the $k$th eigenfunction of the continuum Laplacian operator $\Delta_Q$. In contrast to this classical approach---in which the reference measure $Q$ and resulting Laplacian $\Delta_Q$ are determined a priori---in Laplacian eigenmaps the eigenvectors of the graph Laplacian serve as the basis. These eigenvectors are data-dependent objects, and adapt to the geometry of the unknown design distribution $P$ in a rich manner. For instance, they respect the cluster structure of $P$, meaning that if the density $p$ has multiple connected components, the first few eigenvectors of $L$ will (with high probability) be piecewise constant over each such component \textcolor{blue}{(vonLuxburg 2009)}. On the other hand if $P$ is supported on some low-dimensional manifold, the graph Laplacian eigenvectors concentrate around the eigenfunctions of a manifold Laplace-Beltrami operator, and thus give a principled embedding of potentially high dimensional design points $X_i$ into a lower dimensional space \citep{belkin03a}.

Thus Laplacian eigenmaps is a data-dependent alternative to classical spectral projection estimators. This data-dependency is appealing. Classical spectral projection estimators possess attractive theoretical properties for nonparametric regression---in particular, they are minimax rate-optimal for regression over H\"{o}lder and Sobolev spaces \cite{tsybakov08,johnstone2011,gine16}---but in practice suffer from some serious drawbacks. A basic difficulty is that finding the eigenfunctions of $\Delta_Q$ is in general non-trivial, so that it may not be possible to compute the estimator in the first place. Moreover, spectral projection estimators make sense (and are minimax optimal) only if the reference measure $Q$ is very close or equal to $P$, since otherwise the basis functions are orthogonalized with respect to the wrong measure. For these reasons, such estimators are typically proposed and studied under very restrictive conditions on the design points: for instance, that they are equally spaced fixed grid points, or that they are random but uniformly distributed on the unit cube $[0,1]^d$. There do exist fixes to the issues just raised, for instance using nonparametric least-squares. We will compare Laplacian eigenmaps to nonparametric least squares in more detail later, and for now merely point out that it fundamentally changes the estimator under consideration.  \textcolor{red}{(For Ryan + Siva): I attempted to edit this but still don't like it.}

In contrast, Laplacian eigenmaps directly approximates a spectral projection method: it projects the responses onto an orthogonal basis (eigenvectors of the graph Laplacian) that approximates the smooth eigenfunctions of $\Delta_P$.  Laplacian eigenmaps is perfectly well-defined, and indeed straightforward to compute, when the design $P$ is unknown. This includes the situation where $P$ may be non-uniform, or even concentrated on a low-dimensional manifold. On the other hand, because graph Laplacian eigenvectors depend on the random design points $X_1,\ldots,X_n$ in \textcolor{red}{(non-trivial ways)}, it is substantially more difficult to analyze Laplacian eigenmaps than to analyze classical spectral projection methods. For this reason, the theoretical statistical properties of Laplacian eigenmaps, and in particular its optimality as a method for nonparametric regression with random design, remain poorly understood.

@@

Broadly seeking, our answer is yes. We consider two different data models, one of which assumes the support $\mc{X}$ of the distribution $P$ is a full-dimensional, flat Euclidean domain, and the other of which assumes that $\mc{X}$ is a manifold of small intrinsic dimension $m$. We show that when the regression function $f_0$ is smooth, in the Sobolev sense of having weak derivatives bounded in $\Leb^2(\mc{X})$ norm, Laplacian eigenmaps methods are statistically minimax optimal, for both estimation and goodness-of-fit testing. Our statements hold for different relations between the dimension $d$ (or $m$) and number of derivatives $s$, depending on the problem (estimation or testing), and are summarized in Tables~\ref{tbl:estimation_rates} and~\ref{tbl:testing_rates}.
\begin{table}
	\begin{center}
		\begin{tabular}{p{.2\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Smoothness order & Flat Euclidean (Model~\ref{def:model_flat_euclidean}) & Manifold (Model~\ref{def:model_manifold}) \\
			\hline
			$s \leq 3$ & $\bm{n^{-2s/(2s + d)}}$ & $\bm{n^{-2s/(2s + m)}}$ \\
			$s > 3$  & $\bm{n^{-2s/(2s + d)}}$ & $n^{-6/(6 + m)}$
		\end{tabular}
	\end{center}
	\caption{Summary of estimation rates over Sobolev balls. Bold font marks minimax optimal rates. In each case, rates hold for all $d \in \mathbb{N}$ (under Model~\ref{def:model_flat_euclidean}), and for all $m \in \mathbb{N}, 1 < m < d$ (under Model~\ref{def:model_manifold}). Although we suppress it for simplicity, in all cases when the Laplacian eigenmaps estimator is optimal, the dependence of the error rate on the radius $M$ of the Sobolev ball is also optimal, as long as $n^{-1/2} \lesssim M \lesssim n^{1/d}$.}
	\label{tbl:estimation_rates}
\end{table}

\begin{table}
	\begin{center}
		\begin{tabular}{p{.175\textwidth} p{.175\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Smoothness order & Dimension & Flat Euclidean (Model~\ref{def:model_flat_euclidean}) & Manifold (Model~\ref{def:model_manifold}) \\
			\hline
			\multirow{2}{*}{$s = 1$} & $\dim(\mc{X}) < 4$ & $\bm{n^{-4s/(4s + d)}}$ & $\bm{n^{-4s/(4s + m)}}$ \\
			& $\dim(\mc{X}) \geq 4$ & $\bm{n^{-1/2}}$ & $\bm{n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s = 2$ or $3$} & $\dim(\mc{X}) \leq 4$  & $\bm{n^{-4s/(4s + d)}}$ & $\bm{n^{-4s/(4s + m)}}$ \\
			& $4 <\dim(\mc{X}) < 4s$  & $n^{-2s/(2(s - 1) + d)}$ & $n^{-2s/(2(s - 1) + m)}$\\
			& $\dim(\mc{X}) \geq 4s$ & $\bm{n^{-1/2}}$ & $\bm{n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s > 3$} & $\dim(\mc{X}) \leq 4$ & $\bm{n^{-4s/(4s + d)}}$ & $n^{-12/(12 + d)}$ \\
			& $4 < \dim(\mc{X}) < 4s$ & $n^{-2s/(2(s - 1) + d)}$ & $n^{-6/(4 + m)}$ \\
			& $\dim(\mc{X}) \geq 4s$ & $\bm{n^{-1/2}}$ & $\bm{n^{-1/2}}$ \\
		\end{tabular}
	\end{center}
	\caption{Summary of Laplacian eigenmaps testing rates over Sobolev balls. Bold font marks minimax optimal rates. Rates when $d > 4s$ assume that $f_0 \in L^4(P,M)$. Although we suppress it for simplicity, in all cases when the Laplacian eigenmaps test is optimal, the dependence of the error rate on the radius $M$ of the Sobolev ball is also optimal, as long as $n^{-1/2} \lesssim M \lesssim n^{1/d}$.}
	\label{tbl:testing_rates}
\end{table}

\paragraph{Related Work.}
There is an incredible amount of work on the properties of classical spectral projection methods, which go far beyond their optimality for nonparametric regression. We will not attempt to summarize this work, and instead refer to \citet{wasserman2006,gyorfi2006,tsybakov08,johnstone2011,gine16} and the references therein.

More recent work has considered regression on a fixed graph, where one treats the design points $x_1,\ldots,x_n$ as vertices in a fixed graph $G$, and carries out inference with respect to the function evaluations $(f_0(x_i))_{i = 1}^{n}$ By this point there exists a relatively mature theory describing this setting. Tight upper bounds have been established that certify the optimality of graph-based methods for both nonparametric estimation \citep{wang2016,hutter2016,sadhanala16,sadhanala17,kirichenko2017,kirichenko2018}) and testing (e.g., \citet{sharpnack2010identifying,sharpnack2013b,sharpnack2013,sharpnack2015} over different ``function'' classes (in quotes because these classes really model the $n$-dimensional vector of evaluations) We call particular attention to~\citet{sadhanala16} and~\citet{sharpnack2015}, who analyze the Laplacian eigenmaps estimator and test statistic, respectively. This setting is quite general, because the graph need not be a geometric graph defined on a vertex set which belongs to Euclidean space. On the other hand, in many situations it may be somewhat to unnatural to assume that the design points are a priori fixed, and that the regression function $f_0$ exhibits ``smoothness'' over this fixed design. Instead, it may be more reasonable to adopt the \emph{random design} perspective that we work in, and assume that the regression function $f_0$ exhibits a more classical notion of smoothness.

However, as already mentioned, when the design points are random so too are the graph Laplacian eigenvectors, and grasping their properties is in general non-trivial. For this reason, there has not been much analysis of random design nonparametric regression using Laplacian eigenmaps. \cite{zhou2011} consider the Laplacian eigenmaps estimator, but in the semi-supervised setting, where one additionally observes unlabeled design points $X_{n + 1},\ldots,X_{n + m}$. Their analysis assumed that for a fixed number of labeled samples $n$, the number of unlabeled samples $m$ grows to infinity. In this case the eigenvectors of the graph Laplacian converge to eigenfunctions of a continuum Laplacian, and the analysis of the resulting estimator is identical to that of a classical spectral projection estimator. \cite{lee2016} consider the \emph{diffusion maps} estimator---which uses the eigenvectors of a different normalization of the graph Laplacian $L$---in both the supervised and semi-supervised setups. In the supervised case, they show that the diffusion maps estimator converges to the regression function as the sample size $n \to \infty$, but at a suboptimal rate. As far as we know, there has been no analysis of the test statistic $\wh{T}$ in the random design framework which we study.

There has been much more analysis of the convergence properties of eigenvectors of random graph Laplacians to their continuum limits.  \citet{belkin07,vonluxburg2008,singer2017, garciatrillos18} show that eigenvalue-eigenvector pairs $(\lambda,v)$ of a graph Laplacian converge to eigenvalue-eigenfunction pairs $(\lambda,\psi)$ of a limiting differential or integral operator. \citet{burago2014, shi2015, trillos2019, calder2019, cheng2021} build on these works, by giving finite sample bounds, rates of convergence, and making statements uniform. These results justify our intuition that Laplacian eigenmaps, which projects the responses ${\bf Y}$ onto eigenvectors of a graph Laplacian, is approximating a classical spectral projection method, which projects ${\bf Y}$ onto eigenfunctions of the limiting differential operator $\Delta_P$. In fact, more formally these results imply that for a fixed number of eigenvectors $K > 0$, the Laplacian eigenmaps estimator converges to its classical counterpart as $n \to \infty$. By taking $K \to \infty$ sufficiently slowly with $n$, we can even conclude that Laplacian eigenmaps will be a consistent estimator of $f_0$. Unfortunately, the resulting rates of convergence implied by this approach are severely suboptimal, compared to the minimax rates. Our analysis showing that Laplacian eigenmaps achieves minimax optimal rates of convergence will use some of the results mentioned above, but will overall proceed by a very different route than the one just considered.

Finally, we point out that there are other ways to use neighborhood graphs, and specifically graph Laplacians, for nonparametric regression. For instance, \cite{trillos2020} \textcolor{blue}{(Green 2021)} use the graph Laplacian to induce a penalty over functions $f: \{\bf X\} \to \Reals$. The \emph{Laplacian smoothing} estimator $\wh{f}_{\mathrm{LS}}$ is obtained by minimizing the sum of this penalty with a data-fidelity term,
\begin{equation*}
\wh{f}_{\mathrm{LS}} := \|Y - f\|_n^2 + \lambda \dotp{L_{n,\varepsilon}f}{f}_n.
\end{equation*}
\textcolor{blue}{(Green 2021)} show that the resulting estimator is minimax optimal, but only for $s = 1$ and $d \leq 4$. In contrast, we show in this work that the Laplacian eigenmaps estimator is optimal for all $s$ and $d$.

@@

We now outline the structure of the rest of this paper. In Section~\ref{sec:setup_main_results}, we formally define the regression problem and estimator we consider. We also give some background about minimax regression over Sobolev spaces, and recall the classical spectral projection estimators which achieve minimax rates of convergence. In Sections~\ref{sec:minimax_optimal_laplacian_eigenmaps} and~\ref{sec:manifold_adaptivity}, we establish our main results regarding the optimality of Laplacian eigenmaps. Since thhe Laplacian eigenmaps estimator $\wh{f}$ is defined only at the design points $X_1,\ldots,X_n$, in Section~\ref{sec:out_of_sample} we propose a method for out-of-sample extension of $\wh{f}$, and show that it has optimal out-of-sample error. In Section~\ref{sec:experiments} we examine the empirical behavior of Laplacian eigenmaps, and compare it to some natural competitors for nonparametric regression. We conclude with some discussion in Section~\ref{sec:discussion}.

\paragraph{Notation.}
We frequently refer to various classical function classes. For a domain $\mc{X}$ with volume form $d\mu$, we let $L^2(\mc{X})$ denote the set of functions $f$ for which $\|f\|_{L^2(\mc{X})}^2 := \int f^2 \,d\mu  < \infty$, and equip $L^2(\mc{X})$ with the norm $\|\cdot\|_{L^2(\mc{X})}$. We define $\dotp{f}{g}_P := \int fg\,dP$, and let $L^2(P)$ contain those functions $f$ for which $\|f\|_P^2 := \dotp{f}{f}_P$ is finite. Finally, we let $L^2(P_n)$ consist of those ``functions'' $f: \set{X_1,\ldots,X_n} \to \Reals$ for which the empirical norm $\|f\|_{n}^2 := \frac{1}{n}\sum_{i = 1}^{n} \bigl(f(X_i)\bigr)^2 < \infty$. When there is no chance of confusion, we will sometimes associate functions in $L^2(P_n)$ with vectors in $\Reals^n$, and vice versa. We use $C^k(\mc{X})$ to refer to functions which are $k$ times continuously differentiable in $\mc{X}$, either for some integer $k \geq 1$ or for $k = \infty$. We let $C_c^{\infty}(\mc{X})$ represent those functions in $C^{\infty}(\mc{X})$ which are compactly contained in $\mc{X}$. We write $\partial f/\partial r_i$ for the partial derivative of $f$ in the $i$th standard coordinate of $\Rd$, and use the multi-index notation $D^{\alpha}f := \partial^{|\alpha|}f/\partial^{\alpha_1}x_1\ldots\partial^{\alpha_d}x_d$ for multi-indices $\alpha \in \Reals^m$.

We write $\|\cdot\|$ for Euclidean, and $d_{\mc{X}}(x',x)$ for the geodesic distance between points $x$ and $x'$ on a manifold $\mc{X}$. Then for a given $\delta > 0$, $B(x,\delta)$ is the radius-$\delta$ ball with respect to Euclidean distance, whereas $B_{\mc{X}}(x,\delta)$ is the radius-$\delta$ ball with respect to geodesic distance. Letting $T_x(\mc{X})$ be the tangent space at a point $x \in \mc{X}$, we write $B_m(v,\delta) \subset T_x(\mc{X})$ for the radius-$\delta$ ball centered at $v \in T_x(\mc{X})$.

For sequences $(a_n)$ and $(b_n)$, we use the asymptotic notation $a_n \lesssim b_n$ to mean that there exists a number $C$ such that $a_n \leq C b_n$ for all $n$. We write $a_n \asymp b_n$ when $a_n \lesssim b_n$ and $b_n \lesssim a_n$. On the other hand we write $a_n = o(b_n)$ when $\lim a_n/b_n = 0$, and likewise $a_n = \omega(b_n)$ when $\lim a_n/b_n = \infty$. Finally $a \vee b := \max\{a,b\}$ and $a \wedge b := \min\{a,b\}$.

@@

In this section, we begin by giving a precise definition of our framework, and the Laplacian eigenmaps methods we study. We then review minimax rates for nonparametric regression over Sobolev spaces. We pay special attention to the classical spectral projection methods which achieve these rates, since such methods are closely connected to Laplacian eigenmaps.

\subsection{Nonparametric regression with random design}
\label{sec:regression_laplacian_eigenmaps}

We will operate in the usual setting of nonparametric regression with random design, in which we observe independent random samples $(X_1,Y_1),\ldots,(X_n,Y_n)$. The design points $X_1,\ldots,X_n$ are sampled from a distribution $P$ with support $\mc{X} \subseteq \Rd$, and the responses follow the signal plus noise model
\begin{equation}
\label{eqn:model}
Y_i = f_0(X_i) + w_i,
\end{equation}
with regression function $f_0: \mc{X} \to \Reals$, and $w_i \sim N(0,1)$ independent Gaussian noise. For simplicity we will assume throughout that the noise has unit-variance, but all of our results extend in a straightforward manner to the case where the variance is equal to a known positive value.

@@

\begin{definition}[Flat Euclidean model]
	\label{def:model_flat_euclidean}
	The data $(X_1,Y_1),\ldots,(X_n,Y_n)$ are sampled according to~\eqref{eqn:model}. The support $\mc{X}$ of the design distribution $P$ is an open, connected, and bounded subset of $\Rd$, with Lipschitz boundary. The distribution $P$ admits a Lipschitz density $p$ with respect to the $d$-dimensional Lebesgue measure $\nu$, which is bounded away from $0$ and $\infty$,
	\begin{equation*}
	0 < p_{\min} \leq p(x) \leq p_{\max} < \infty, \quad \textrm{for all $x \in \mc{X}$.}
	\end{equation*}
\end{definition}
